Recovery of a deleted SARS-CoV-2 deep sequencing data set from the early Wuhan epidemic

### 2021-06-15 paper/paper.tex
more writing of Results section existing sequences

The SRA is designed as a permanent archive of deep sequencing data.
The SRA documentation states that after a sequencing run is uploaded, ``neither its files can be replaced nor filenames can be changed,'' and that data can only be deleted by e-mailing SRA staff~\citep{SRA_deletion}.
An example of this process from another study is in Figure~\ref{fig:pangolin_emails}, which shows an e-mail sent to SRA staff by the lead author of a paper on pangolin coronaviruses~\citep{xiao2020isolation} to request the deletion of two sequencing runs.
Therefore, subsequent to March 30, 2020, a similar e-mail request must have been made to fully delete SARS-CoV-2 deep sequencing project PRJNA612766.

@@

The pre-print~\citep{wang2020medRxiv} says the approach was applied to ``45 nasopharyngeal swab samples from outpatients with suspected COVID-19 early in the epidemic.''
The digital object identifier (DOI) for the pre-print indicates that it was processed by \textit{medRxiv} on March 4, 2020, which is one day after China's State Council ordered that all papers related to COVID-19 must be approved by a centralized group to ensure that scientific publications were coordinated ``like moves in a game of chess''~\citep{Kang2020}.
The final published manuscript~\citep{wang2020small} from June of 2020 updated the description from ``early in the epidemic'' to ``early in the epidemic (January 2020).''
Both the pre-print and published manuscript say that 34 of the 45 early epidemic samples were positive in the sequencing-based diagnostic approach.
In addition, both state that the approach was later applied to 16 additional samples collected on February 11 and 12 of 2020 from SARS-CoV-2 patients hospitalized at Renmin Hospital of Wuhan University.

@@

\begin{figure*}[]
\centering
\includegraphics[width=\linewidth]{figures/coverage_region.pdf}
\caption{Sequencing depth over the SARS-CoV-2 genome from site 21,570 to 29,550 for the 34 virus-positive early outbreak samples and the 16 samples from hospitalized patients in February.
Depth is the number of aligned reads that cover that site with a quality score $\ge$20.
The dashed gray line is the minimum coverage required to call a consensus identity at a site.
Note that the y-axis uses a symlog scale.
An interactive version of this plot that enables zooming into specific site ranges and mouseovers to see read count statistics at each site is at {\color{red} add link}.
A version of the plot where the x-axis spans the entire SARS-CoV-2 genome is in Supplementary Figure~S\ref{suppfig:coverage_all}.
}
\label{fig:coverage}
\end{figure*}

@@

I called the sequence identity at each site with coverage $\ge$3 and $>$80\% of the reads concurring on the nucleotide identity.
With these criteria, 13 of the early outpatient samples and 1 of the February hospitalized patient samples had sufficient coverage to call the nucleotide sequence at $>$95\% of the sites in the region of interest (Table~\ref{tab:mutations}), and for the remainder of this paper I focus on these high-coverage samples.
Table~\ref{tab:mutations} also shows the mutations in each sample relative to proCoV2, which is a putative progenitor of SARS-CoV-2 inferred by \citet{kumar2021evolutionary} that differs from the widely using Wuhan-Hu-1 reference sequence by three mutations (C8782T, C18060T, and T28144C).
Although requiring a coverage of only $\ge$3 is relatively lenient, Table~\ref{tab:mutations} shows that all sites with mutations are supported by at least 10 reads.
In addition, the mutations called from the raw sequences in Table~\ref{tab:mutations} concord with those mentioned in \citet{wang2020small}.

\subsection{Phylogenetic analysis of existing SARS-CoV-2 sequences from January of 2020 or earlier}
To contextualize the sequences recovered from the deleted project described above, here I first analyze early SARS-CoV-2 sequences already available in the GISAID database~\citep{shu2017gisaid}.

Numerous evolutionary analyses have shown that known human SARS-CoV-2 sequences are consistent with expansion from a single progenitor sequence~\citep{kumar2021evolutionary,pekar2021timing,rambaut2020dynamic,forster2020phylogenetic,pipes2021assessing}.
However, attempts to infer the identity of the SARS-CoV-2 progenitor sequence have been confounded by a perplexing fact: the sequences of the earliest reported isolates from Wuhan are \emph{not} the sequences most similar to SARS-CoV-2's bat coronavirus relatives~\citep{pipes2021assessing}.

\begin{figure*}
\centering
\includegraphics[width=\linewidth]{figures/deltadist_RaTG13.pdf}
\caption{The reported collection date of each SARS-CoV-2 sequence versus its relative mutational distance from the RaTG13 bat coronavirus outgroup.
Mutational distances are relative to the putative progenitor proCoV2 inferred by \citet{kumar2021evolutionary}.
The plot shows all sequences in GISAID collected no later than February 28, 2020, after performing the quality-control described in the Methods.
Sequences that the joint WHO-China report~\cite{CITE} describes as being associated with the Wuhan Seafood Market are plotted with squares.
Points are slightly jittered on the y-axis.
Go to {\color{red} add link} for an interactive version of this plot that enables toggling of the outgroup to RpYN06 and RmYN02, mouseovers to see details for each point including strain name and mutations relative to proCoV2, and adjustment of the y-axis jittering.
Static versions of the plot with RpYN06 and RmYN02 outgroups are in Supplementary Figure~\ref{suppfig:deltadist_RpYN06_RmYN02}.
}
\label{fig:deltadist_RaTG13}
\end{figure*}

>>>>

The SRA is designed as a permanent archive of deep sequencing data.
The SRA documentation states that after a sequencing run is uploaded, ``neither its files can be replaced nor filenames can be changed,'' and that data can only be deleted by e-mailing SRA staff~\citep{SRA_deletion}.
An example of this process from another study is in Figure~\ref{fig:pangolin_emails}, which shows an e-mail sent to SRA staff by the lead author of a paper on pangolin coronaviruses~\citep{xiao2020isolation} to request the deletion of two sequencing runs.
Subsequent to March 30, 2020, a similar e-mail request must have been made to fully delete SARS-CoV-2 deep sequencing project PRJNA612766.

@@

The pre-print~\citep{wang2020medRxiv} says the approach was applied to ``45 nasopharyngeal swab samples from outpatients with suspected COVID-19 early in the epidemic.''
The digital object identifier (DOI) for the pre-print indicates that it was processed by \textit{medRxiv} on March 4, 2020, which is one day after China's State Council ordered that all papers related to COVID-19 must be approved by a centralized group to ensure that scientific publications were coordinated ``like moves in a game of chess''~\citep{Kang2020}.
The final published manuscript~\citep{wang2020small} from June of 2020 updated the description from ``early in the epidemic'' to ``early in the epidemic (January 2020).''
Both the pre-print and published manuscript say that 34 of the 45 early epidemic samples were positive in the sequencing-based diagnostic approach.
In addition, both state that the approach was later applied to 16 additional samples collected on February 11--12, 2020, from SARS-CoV-2 patients hospitalized at Renmin Hospital of Wuhan University.

@@

\begin{figure*}[]
\centering
\includegraphics[width=\linewidth]{figures/coverage_region.pdf}
\caption{Sequencing depth over the SARS-CoV-2 genome from site 21,570 to 29,550 for the 34 virus-positive early epidemic samples and the 16 samples from hospitalized patients in February.
Depth is the number of aligned reads that cover that site with a quality score $\ge$20.
The dashed gray line is the minimum coverage required to call a consensus identity at a site.
Note that the y-axis uses a symlog scale.
An interactive version of this plot that enables zooming into specific site ranges and mouseovers to see read count statistics at each site is at {\color{red} add link}.
A version of the plot where the x-axis spans the entire SARS-CoV-2 genome is in Supplementary Figure~S\ref{suppfig:coverage_all}.
}
\label{fig:coverage}
\end{figure*}

@@

I called the consensus viral sequence for each sample at each site with coverage $\ge$3 and $>$80\% of the reads concurring on the nucleotide identity.
With these criteria, 13 of the early outpatient samples and 1 of the February hospitalized patient samples had sufficient coverage to call the consensus sequence at $>$95\% of the sites in the region of interest (Table~\ref{tab:mutations}), and for the remainder of this paper I focus on these high-coverage samples.
Table~\ref{tab:mutations} also shows the mutations in each sample relative to proCoV2, which is a putative progenitor of SARS-CoV-2 inferred by \citet{kumar2021evolutionary} that differs from the widely using Wuhan-Hu-1 reference sequence by three mutations (C8782T, C18060T, and T28144C).
Although requiring coverage of only $\ge$3 is relatively lenient, Table~\ref{tab:mutations} shows that all sites with mutations have coverage $\ge$10.
In addition, the mutations I called from the raw sequence data in Table~\ref{tab:mutations} concord with those mentioned in \citet{wang2020small}.

\subsection{Analysis of existing sequences emphasizes the perplexing discordance between collection date and distance to SARS-CoV-2's bat coronavirus relatives}
To contextualize the viral sequences recovered from the deleted project described above, here I first analyze early SARS-CoV-2 sequences already available in the GISAID database~\citep{shu2017gisaid}.

Evolutionary analyses have shown that known human SARS-CoV-2 sequences are consistent with expansion from a single progenitor sequence~\citep{kumar2021evolutionary,pekar2021timing,rambaut2020dynamic,forster2020phylogenetic,pipes2021assessing}.
However, attempts to infer this SARS-CoV-2 progenitor sequence have been confounded by a perplexing fact: the earliest reported sequences from Wuhan are \emph{not} the sequences most similar to SARS-CoV-2's bat coronavirus relatives~\citep{pipes2021assessing}.
This fact is perplexing because although the proximal origin of SARS-CoV-2 remains unclear (i.e., zoonosis versus lab accident), all reasonable explanations agree that at a deeper level the SARS-CoV-2 genome is derived from bat coronaviruses~\citep{lytras2021exploring}.
One would therefore expect the first reported SARS-CoV-2 sequences to be the most similar to the virus's bat coronavirus relatives---but this is not the case.

\begin{figure*}
\centering
\includegraphics[width=\linewidth]{figures/deltadist_RaTG13.pdf}
\caption{The reported collection date of each SARS-CoV-2 sequence versus its relative mutational distance from the RaTG13 bat coronavirus outgroup.
Mutational distances are relative to the putative progenitor proCoV2 inferred by \citet{kumar2021evolutionary}.
The plot shows all sequences in GISAID collected no later than February 28, 2020, after performing the quality-control described in the Methods.
Sequences that the joint WHO-China report~\citep{WHO2021origins} describes as being associated with the Wuhan Seafood Market are plotted with squares.
Points are slightly jittered on the y-axis.
Go to {\color{red} add link} for an interactive version of this plot that enables toggling of the outgroup to RpYN06 and RmYN02, mouseovers to see details for each point including strain name and mutations relative to proCoV2, and adjustment of the y-axis jittering.
Static versions of the plot with RpYN06 and RmYN02 outgroups are in Supplementary Figure~\ref{suppfig:deltadist_RpYN06_RmYN02}.
}
\label{fig:deltadist_RaTG13}
\end{figure*}

This conundrum is illustrated in Figure~\ref{fig:deltadist_RaTG13}, which plots the collection date of SARS-CoV-2 sequences in GISAID versus the relative number of mutational differences from RaTG13~\citep{zhou2020pneumonia}, which is the bat coronavirus with the highest overall identity to SARS-CoV-2.
The earliest SARS-CoV-2 sequences were collected in Wuhan in December, but these sequences are more distant from RaTG13 than SARS-CoV-2 sequences collected in January from other locations in China or even other countries (Figure~\ref{fig:deltadist_RaTG13}).
The discrepancy is especially pronounced for early Wuhan sequences collected from patients who had visited the Huanan Seafood Market, which was widely reported as an initial location of the outbreak and has been postulated to be a site of potential zoonosis~\citep{WHO2021origins}.
All sequences associated with the Huanan Seafood Market differ from RaTG13 by at least three more mutations than sequences subsequently collected at various other locations (Figure~\ref{fig:deltadist_RaTG13})---a fact that is difficult to reconcile with the idea that the market was the original site of spread of a bat coronavirus into humans.
Importantly, all these observations also hold true if SARS-CoV-2 is compared to other related bat coronaviruses~\citep{lytras2021exploring} such as RpYN06~\citep{zhou2021identification} or RmYN02~\citep{zhou2020novel} rather than RaTG13 (Supplementary Figure~S\ref{suppfig:deltadist_RpYN06_RmYN02}).

This conundrum can also be visualized in a phylogenetic context by rooting a tree of early SARS-CoV-2 sequences so that the ``progenitor'' sequence is chosen to be the one closest to the bat coronavirus outgroup.
If we limit the analysis to sequences with at least two observation among strains collected no later than January 2020, there are three possible ways to root the tree 
 (Figure~\ref{fig:tree_RaTG13} shows these rootings using a RaTG13 outgroup; Supplementary Figure~S\ref{suppfig:tree_RpYN06_RmYN02} shows identical results are obtained using RpYN06 or RmYN02 outgroups).
